---
title: "Introduction to Rvoterdistance"
author: "Loren Collingwood"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Rvoterdistance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**Rvoterdistance** calculates the geographic distance between voters and
polling locations (or vote-by-mail drop boxes) using the Haversine
great-circle formula, implemented in C++ for speed. The package supports:

- **Nearest location**: find the single closest polling place for each voter
- **k-nearest locations**: find the k closest locations per voter
- **Distance threshold**: find all locations within a specified radius
- **Boundary distance**: find how far each voter lives from a geographic
  boundary (state border, river, district line, etc.)
- **sf integration**: pass `sf` POINT geometries directly

## Installation

```{r install, eval = FALSE}
# From GitHub:
remotes::install_github("lorenc5/Rvoterdistance")
# or
remotes::install_github("RPVote/Rvoterdistance")
```

## Included Data

The package ships with two example datasets:

- `king_dbox`: King County, WA ballot drop box locations and a sample of
  voters
- `meck_ev`: Mecklenburg County, NC early voting locations and a sample of
  voters

```{r data}
library(Rvoterdistance)
data(meck_ev)

str(voter_meck)
str(early_meck)
```

## Basic Usage: Nearest Location

The main function is `nearest_location()`. With the default `k = 1`, it
returns one row per voter with the distance to the nearest polling
location:

```{r nearest}
result <- nearest_location(
  voters = voter_meck,
  locations = early_meck,
  voter_coords = c("lat", "long"),
  location_coords = c("lat", "long")
)

head(result)
```

The output includes the voter data, the matched location data, and three
distance columns: `distance_m` (meters), `distance_km`, and
`distance_miles`.

## k-Nearest Locations

To find the 3 closest early voting sites for each voter:

```{r knearest}
result_k3 <- nearest_location(
  voter_meck, early_meck,
  voter_coords = c("lat", "long"),
  location_coords = c("lat", "long"),
  k = 3,
  append_data = FALSE
)

head(result_k3, 9)
```

The output is in long format with a `rank` column (1 = nearest).

## Distance Threshold

Find all early voting locations within 5 miles of each voter:

```{r threshold}
result_5mi <- nearest_location(
  voter_meck[1:20, ], early_meck,
  voter_coords = c("lat", "long"),
  location_coords = c("lat", "long"),
  max_dist = 5,
  units = "miles",
  append_data = FALSE
)

head(result_5mi, 10)

# How many locations within 5 miles per voter?
table(result_5mi$voter_id)
```

## Using sf Objects

If your data are already `sf` POINT objects, pass them directly ---
no need to specify coordinate column names:

```{r sf, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

voters_sf <- st_as_sf(voter_meck, coords = c("long", "lat"), crs = 4326)
locs_sf <- st_as_sf(early_meck, coords = c("long", "lat"), crs = 4326)

result_sf <- nearest_location(voters_sf, locs_sf, append_data = FALSE)
head(result_sf)
```

If the CRS is not WGS-84 (EPSG:4326), the package automatically
transforms to WGS-84 and prints a message.

## Convenience Functions

For quick calculations without the full `nearest_location()` interface:

```{r convenience}
# Minimum distance in km for each voter
km <- dist_km(
  voter_meck$lat, voter_meck$long,
  early_meck$lat, early_meck$long
)
summary(km)

# Minimum distance in miles
mi <- dist_mile(
  voter_meck$lat, voter_meck$long,
  early_meck$lat, early_meck$long
)
summary(mi)

# Single-pair distance (e.g., Charlotte to Raleigh)
haversine(35.2271, -80.8431, 35.7796, -78.6382, units = "miles")
```

## Distance to a Geographic Boundary

The `dist_to_boundary()` function computes the minimum distance from
each voter to a geographic boundary such as a state border, river, or
district line. The boundary is provided as an `sf` geometry object
(LINESTRING, MULTILINESTRING, POLYGON, or MULTIPOLYGON). The
computation uses the spherical cross-track distance formula in C++ with
bounding-box pruning, making it practical for large voter files.

### Simple boundary line

```{r boundary-line, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Simplified AZ-NM border: a vertical line at longitude -109.05
border <- st_sf(
  geometry = st_sfc(
    st_linestring(matrix(c(-109.05, 31.33,
                            -109.05, 37.00),
                          ncol = 2, byrow = TRUE)),
    crs = 4326
  )
)

# Two voters: one in Albuquerque, one near the border
voters <- data.frame(
  name = c("Albuquerque voter", "Border voter"),
  lat  = c(35.08, 35.0),
  lon  = c(-106.65, -109.0)
)

d <- dist_to_boundary(voters, border,
  voter_coords = c("lat", "lon"),
  units = "km", progress = FALSE
)

data.frame(voter = voters$name, dist_km = round(d, 1))
```

### Polygon boundary

When the boundary is a polygon, `dist_to_boundary()` measures the
distance to the polygon's **perimeter** (nearest edge), not to its
interior. A point inside the polygon returns the positive distance to
the nearest edge.

```{r boundary-polygon, eval = requireNamespace("sf", quietly = TRUE)}
# A rectangular district
district <- st_sf(
  geometry = st_sfc(
    st_polygon(list(matrix(c(
      -110, 35,
      -108, 35,
      -108, 37,
      -110, 37,
      -110, 35
    ), ncol = 2, byrow = TRUE))),
    crs = 4326
  )
)

# One voter inside, one outside
voters2 <- data.frame(
  name = c("Inside district", "Outside district"),
  lat  = c(36.0, 36.0),
  lon  = c(-109.0, -107.0)
)

d2 <- dist_to_boundary(voters2, district,
  voter_coords = c("lat", "lon"),
  units = "miles", progress = FALSE
)

data.frame(voter = voters2$name, dist_miles = round(d2, 1))
```

### Using sf POINT voters

If your voter data is already an `sf` object with POINT geometry, pass
it directly --- no need for `voter_coords`:

```{r boundary-sf-voters, eval = requireNamespace("sf", quietly = TRUE)}
voters_sf <- st_sf(
  name = c("Voter A", "Voter B"),
  geometry = st_sfc(
    st_point(c(-106.65, 35.08)),
    st_point(c(-109.00, 35.00)),
    crs = 4326
  )
)

d3 <- dist_to_boundary(voters_sf, border,
  units = "miles", progress = FALSE
)

data.frame(voter = voters_sf$name, dist_miles = round(d3, 1))
```

### Supported units

`dist_to_boundary()` supports `"km"` (default), `"miles"`, and
`"meters"`:

```{r boundary-units, eval = requireNamespace("sf", quietly = TRUE)}
voters_u <- data.frame(lat = 36.0, lon = -108.0)

d_km <- dist_to_boundary(voters_u, border,
  voter_coords = c("lat", "lon"), units = "km", progress = FALSE)
d_mi <- dist_to_boundary(voters_u, border,
  voter_coords = c("lat", "lon"), units = "miles", progress = FALSE)
d_m  <- dist_to_boundary(voters_u, border,
  voter_coords = c("lat", "lon"), units = "meters", progress = FALSE)

data.frame(km = round(d_km, 2), miles = round(d_mi, 2), meters = round(d_m, 1))
```

## Performance

The Haversine computation runs in C++ and uses partial sorting
(`std::nth_element`) for k-nearest queries, giving O(n) per voter
instead of O(n log n). The `dist_to_boundary()` function uses
bounding-box pruning to skip distant boundary segments, avoiding
unnecessary cross-track distance calculations. For large voter files,
enable progress reporting:

```{r progress, eval = FALSE}
result <- nearest_location(
  big_voter_file, locations,
  voter_coords = c("lat", "lon"),
  location_coords = c("lat", "lon"),
  k = 3,
  progress = TRUE
)

# Boundary distance with progress
d <- dist_to_boundary(
  big_voter_file, state_border,
  voter_coords = c("lat", "lon"),
  progress = TRUE
)
```

## Application: Geographic Regression Discontinuity

A natural application of `dist_to_boundary()` is a **geographic
regression discontinuity design (RDD)**. The distance to a boundary
serves as the running variable (score), with the boundary itself as
the cutoff. Voters on one side receive a "treatment" (e.g., different
jurisdiction, policy environment, or services) and voters on the other
side serve as controls.

This example simulates 50,000 voters around the Sandia Pueblo
reservation in New Mexico and estimates the effect of reservation
residence on voter turnout using the `rdrobust` package.

```{r rdd-check, include = FALSE}
has_rdd_pkgs <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("rdrobust", quietly = TRUE)
```

### Step 1: Define the reservation boundary

We use a simplified polygon approximating the Sandia Pueblo
reservation, which sits north of Albuquerque between the Rio Grande
and the Sandia Mountains.

```{r rdd-boundary, eval = has_rdd_pkgs}
library(sf)
library(ggplot2)
library(rdrobust)

# Simplified Sandia Pueblo reservation boundary
sandia_coords <- matrix(c(
  -106.6140, 35.1850,
  -106.5400, 35.1800,
  -106.4800, 35.2000,
  -106.4500, 35.2350,
  -106.4450, 35.2700,
  -106.4600, 35.3050,
  -106.4900, 35.3200,
  -106.5500, 35.3250,
  -106.5900, 35.3100,
  -106.6100, 35.2750,
  -106.6200, 35.2350,
  -106.6140, 35.1850   # close the ring
), ncol = 2, byrow = TRUE)

sandia <- st_sf(
  name = "Sandia Pueblo",
  geometry = st_sfc(st_polygon(list(sandia_coords)), crs = 4326)
)
```

### Step 2: Simulate 50,000 voters

Voters are drawn from a mixture of clusters reflecting actual
population centers around the reservation: northeast Albuquerque
(largest cluster), Bernalillo, Corrales, Rio Rancho, Placitas, and
a sparser set on the reservation itself.

```{r rdd-simulate, eval = has_rdd_pkgs}
set.seed(2024)
n <- 50000

# Population centers: lat, lon, mixture weight, spatial spread
#   1. NE Albuquerque / Sandia Heights  (south of reservation, dense)
#   2. Bernalillo                        (northwest, medium)
#   3. On/near reservation               (sparse)
#   4. Rio Rancho                        (west, medium)
#   5. Placitas                          (northeast, small)
#   6. Corrales                          (west along the river)
centers <- data.frame(
  lat    = c(35.160, 35.310, 35.250, 35.275, 35.340, 35.235),
  lon    = c(-106.520, -106.560, -106.530, -106.680, -106.440, -106.630),
  weight = c(0.35, 0.18, 0.12, 0.17, 0.06, 0.12),
  sd_lat = c(0.035, 0.025, 0.035, 0.025, 0.018, 0.018),
  sd_lon = c(0.035, 0.025, 0.040, 0.025, 0.018, 0.018)
)

cluster <- sample(nrow(centers), n, replace = TRUE, prob = centers$weight)
voters <- data.frame(
  voter_id = seq_len(n),
  lat = rnorm(n, mean = centers$lat[cluster], sd = centers$sd_lat[cluster]),
  lon = rnorm(n, mean = centers$lon[cluster], sd = centers$sd_lon[cluster])
)
```

### Step 3: Compute the score variable

The score is the signed distance (in miles) from each voter to the
reservation boundary: **positive** for voters inside the reservation,
**negative** for voters outside, with **zero** at the boundary.

```{r rdd-score, eval = has_rdd_pkgs}
# Distance to the reservation boundary (unsigned, in miles)
dist_miles <- dist_to_boundary(
  voters, sandia,
  voter_coords = c("lat", "lon"),
  units = "miles",
  progress = FALSE
)

# Determine which voters fall inside the reservation
voters_sf <- st_as_sf(voters, coords = c("lon", "lat"), crs = 4326)
inside <- lengths(st_intersects(voters_sf, sandia)) > 0

# Signed score: positive inside, negative outside
voters$score <- ifelse(inside, dist_miles, -dist_miles)
voters$inside <- inside

cat("Voters inside reservation:", sum(inside), "\n")
cat("Voters outside reservation:", sum(!inside), "\n")
summary(voters$score)
```

### Step 4: Generate voter turnout

We simulate a binary turnout variable where the probability of
voting is higher outside the reservation than inside, with a
discontinuous jump at the boundary. A mild gradient in distance
adds realism.

```{r rdd-turnout, eval = has_rdd_pkgs}
# Turnout probability: ~62% far outside, dropping to ~45% just inside
turnout_prob <- 0.62 + 0.005 * voters$score   # gentle gradient
turnout_prob[inside] <- turnout_prob[inside] - 0.15  # discontinuity
turnout_prob <- pmin(pmax(turnout_prob, 0.10), 0.90)  # clamp

voters$voted <- rbinom(n, 1, turnout_prob)

cat("Overall turnout rate:", round(mean(voters$voted), 3), "\n")
cat("Turnout inside:     ", round(mean(voters$voted[inside]), 3), "\n")
cat("Turnout outside:    ", round(mean(voters$voted[!inside]), 3), "\n")
```

### Step 5: Histogram of the score variable

The score distribution shows voter density on each side of the
boundary. The reservation interior (positive scores) is sparsely
populated relative to the surrounding communities.

```{r rdd-histogram, eval = has_rdd_pkgs, fig.width = 7, fig.height = 4}
ggplot(voters, aes(x = score)) +
  geom_histogram(
    aes(fill = inside),
    bins = 100, alpha = 0.8, boundary = 0
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("Outside reservation", "Inside reservation"),
    name = NULL
  ) +
  labs(
    x = "Distance to reservation boundary (miles)",
    y = "Number of voters",
    title = "Score variable: signed distance to Sandia Pueblo boundary"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Step 6: Map of voters and reservation

```{r rdd-map, eval = has_rdd_pkgs, fig.width = 7, fig.height = 6}
# Plot a random sample of 5,000 voters for readability
set.seed(99)
voter_sample <- voters[sample(n, 5000), ]

ggplot() +
  geom_point(
    data = voter_sample,
    aes(x = lon, y = lat, color = score),
    size = 0.4, alpha = 0.6
  ) +
  geom_sf(
    data = sandia,
    fill = NA, color = "black", linewidth = 1
  ) +
  scale_color_gradient2(
    low = "steelblue", mid = "grey90", high = "firebrick",
    midpoint = 0, name = "Score\n(miles)"
  ) +
  labs(
    x = "Longitude", y = "Latitude",
    title = "Simulated voters around Sandia Pueblo reservation",
    subtitle = "Red = inside reservation, Blue = outside"
  ) +
  coord_sf(
    xlim = c(-106.78, -106.35),
    ylim = c(35.08, 35.42)
  ) +
  theme_minimal()
```

### Step 7: RD estimation with `rdrobust`

We estimate the local average treatment effect at the boundary using
`rdrobust()`. The running variable is the signed distance score and
the cutoff is zero.

```{r rdd-estimate, eval = has_rdd_pkgs}
rd <- rdrobust(y = voters$voted, x = voters$score, c = 0)
summary(rd)
```

The estimated coefficient represents the discontinuous change in
turnout probability at the reservation boundary. A negative estimate
indicates lower turnout just inside the reservation relative to just
outside.

### Step 8: RD plot

The `rdplot()` function visualizes the local polynomial fit on each
side of the cutoff, with binned means showing the underlying data
pattern.

```{r rdd-plot, eval = has_rdd_pkgs, fig.width = 7, fig.height = 5}
rdplot(
  y = voters$voted,
  x = voters$score,
  c = 0,
  title = "Geographic RD: Voter turnout at Sandia Pueblo boundary",
  x.label = "Distance to reservation boundary (miles)",
  y.label = "Voter turnout"
)
```
